---
title: Roundup of growth chart packages
author: Monica Gerber
date: '2018-07-01'
categories:
  - conferences
  - roundup
  - ropensci
tags:
  - agd
  - bmi
  - childsds
  - growthchart
  - growthstandards
  - zscorer
slug: roundup-of-growth-chart-packages
---

```{r include=FALSE}
r_ladies_theme <- function(){
  theme_bw() %+replace% 
    theme(text = element_text(family = "HelveticaNeue", face = "plain",
                              colour = 'black', size = 10,
                              hjust = .5, vjust = .5, angle = 0, 
                              lineheight = 1.1, 
                              margin = margin(t = 0, r = 0, b = 0, l = 0, 
                                              unit = "pt"), 
                              debug= FALSE), 
          axis.text = element_text(colour = "#181818"), 
          axis.title = element_text(face = "bold", colour = "#88398A", size = rel(1.1)), 
          plot.title = element_text(face = "bold", size = rel(1.4), 
                                    colour = "#88398A", 
                                    margin = margin(t = 0, r = 0, b = 6.6,
                                                    l = 0, unit = "pt"),
                                    hjust = 0.0),
          legend.title = element_text(face = "bold", colour = "#181818"),
          panel.grid.major = element_line(color = "#D3D3D3"))
}
```


At rOpenSci's annual [unconf](http://unconf18.ropensci.org/), I [suggested](https://github.com/ropensci/unconf18/issues/49) a project to work on functions that convert physical growth measurements into z-scores and percentiles. For example, researchers in the United States studying childhood obesity often calculate the BMI z-score or percentage of the 95th percentile of BMI from nationally [representative survey data](https://www.cdc.gov/growthcharts/cdc_charts.htm).

During the process of brainstorming and working on this project, the team[^1] I was working with found out some related R packages that do this. To help us move forward with the project and identify next steps, I round up these packages below.[^2] If I'm missing something, please comment so I can add it to this post!

# Anthropometric measurements

The [National Health and Nutrition Examination Survey](https://www.cdc.gov/nchs/nhanes/index.htm) (NHANES) has data on height, weight, blood pressure, and other measurements for adolescents from 0 to 239 months that I will use to explore these packages. I'm converting the original variables names to names that are easier to use. The data is in an [.xpt](https://www.loc.gov/preservation/digital/formats/fdd/fdd000464.shtml) file, which I'll read in using the `haven` package.

```{r warning=FALSE}
library(tidyverse)
library(haven)

demo_data <- read_xpt(url("https://wwwn.cdc.gov/Nchs/Nhanes/2015-2016/DEMO_I.XPT"))

demo_need <- demo_data %>%
  select(id = SEQN, 
         agemos = RIDEXAGM, 
         sex_num = RIAGENDR)

exam_data <- read_xpt(url("https://wwwn.cdc.gov/Nchs/Nhanes/2015-2016/BMX_I.XPT"))

exam_need <- exam_data %>%
  select(id = SEQN, 
         bmi = BMXBMI, 
         height_cm = BMXHT, 
         length_cm = BMXRECUM, 
         weight_kg = BMXWT)

df <- demo_need %>%
  left_join(exam_need, by = "id") %>%
  filter(!is.na(agemos)) %>%
  mutate(sex_mf = if_else(sex_num == 1, "M", "F"),
         sex = if_else(sex_num == 1, "Male", "Female"),
         # recumbent length is used if under 2 years old
         height_length_cm = if_else(agemos < 24, length_cm, height_cm),
         ageyears = agemos/12)

# xpt file had column labels, remove column labels 
child_df <- df[1:nrow(df), ]
```

# AGD package

The Analysis of Growth Data (AGD) package is developed by Stef van Buuren. More info [here](https://stefvanbuuren.github.io/AGD/). 

> Tools for the analysis of growth data: to extract an LMS table from a gamlss object, to calculate the standard deviation scores and its inverse, and to superpose two wormplots from different models. The package contains a some varieties of reference tables, especially for The Netherlands.

It has functions for reference tables from the CDC 2000 growth charts, Third Dutch Growth Study (1980), Fourth Dutch Growth Study (1997), and WHO growth charts. The function `y2z()` converts measurements into standard deviation scores (z-scores) and `z2y()` does the opposite, convert standard deviation scores into measurements.

The package is available on CRAN.
```{r eval=FALSE}
install.packages("AGD")
```

To calculate the age- and sex-adjusted BMI z-scores for the children in the NHANES dataset, we can do the following:

```{r}
library(AGD)

bmi_levels <- c("Underweight", "Healthy weight", "Overweight", 
                "Obesity", "Severe obesity")

agd_df <- child_df %>%
  # the CDC growth charts are typically used for children 2 years or older
  filter(agemos >= 24) %>%
  mutate(bmiz =  AGD::y2z(y = bmi, x = ageyears, sex = sex_mf, ref = cdc.bmi),
         bmipct = pnorm(bmiz),
         # what is the 95th percentile for the child's age and sex?,
         z = qnorm(.95),
         p95 = AGD::z2y(z = z, x= ageyears, sex = sex_mf, ref = cdc.bmi),
         # percentage of the 95th percentile
         bmipct95 = bmi/p95,
         # bmi category
         bmicat = factor(case_when(bmipct < .05 ~ "Underweight",
                                   bmipct < .85 ~ "Healthy weight",
                                   bmipct < .95 ~ "Overweight",
                                   bmipct95 < 1.2 ~ "Obesity",
                                   bmipct95 >= 1.2 ~ "Severe obesity"),
                         levels = bmi_levels))

agd_df %>%
  select(agemos, bmi, bmiz, bmipct, bmipct95) %>%
  skimr::skim()
```

A plot of the category values:

```{r}
agd_df %>%
  filter(!is.na(bmicat)) %>%
  ggplot() +
  stat_count(mapping = aes(x = bmicat, y=..prop.., group=1)) +
  r_ladies_theme() +
  labs(title = "BMI categories for U.S. children 2 years and older",
       subtitle = "NHANES 2015-2016; Calculated using the AGD package",
       y = "Proportion",
       x = "BMI Category")
```

*Note: I haven't taken the complex survey design into account here, so these proportions are not an accurate estimate of the prevalence in the U.S. See [this report](https://www.cdc.gov/nchs/data/databriefs/db288.pdf) for those estimates.*

# growthstandards / hgbd

The [growthstandards](https://github.com/hbgdki/growthstandards) package, developed by Ryan Hafen, is an offshoot from a larger package, hbgd, and some docs for the growth standards calculations can be found [here](http://hbgdki.github.io/hbgd/#who_growth_standard). It provides calculations for:

* WHO height, weight, head circumference, muacc, subscalpular skinfold, triceps skinfold
* INTERGROWTH fetal head circumference, biparietel diameter, occipito-frontal diameter, abdominal circumference, femur length
* INTERGROWTH newborn (including very preterm) weight, length, and head circumference

The package has some very useful functions for converting units such as:

* `months2days()` for converting age in months to age in days, and
* `lb2kg()` for pounds to kilograms.

```{r eval=FALSE}
devtools::install_github("HBGDki/growthstandards")
```

The Center for Disease Control (CDC) and American Academy of Pediatrics (AAP) [recommend](https://www.cdc.gov/nccdphp/dnpao/growthcharts/who/recommendations/index.htm) using the WHO growth standards for children in the United States under 2 years old. To calculate the age- and sex-adjusted BMI z-scores for the children in the NHANES dataset using the WHO reference charts we can do the following:

```{r}
child_who <- child_df %>%
  filter(agemos < 24) %>%
  mutate(
    bmi = weight_kg/((length_cm/100)^2) ,
    agedays = growthstandards::months2days(agemos),
    # measurements to bmi z-score
    bmiz = growthstandards::who_bmi2zscore(
      agedays = agedays,
      bmi = bmi,
      sex = sex),
    # measurements to percentiles
    bmipct = growthstandards::who_bmi2centile(
      agedays = agedays, 
      bmi = bmi, 
      sex = sex),
    # find the 95th percentile for age and sex
    p95 = growthstandards::who_centile2bmi(
      agedays = agedays, 
      p = 95, 
      sex = sex),
    # calculate the percentage of the 95th
    bmipct95 = (bmi/p95)*100,
    # obesity?
    obesity = if_else(bmipct >= 95, 1, 0))


child_who %>%
  select(bmi, bmiz, bmipct, bmipct95, obesity) %>%
  skimr::skim()
```

Density plots of our new values:

```{r}
vars <- c("bmi", "bmiz", "bmipct", "bmipct95")
sub <- "Using WHO growth charts from growthstandards package"

density_fun <- function(xvar, df, sub) {
  myplot <- ggplot(df, aes_string(x = xvar)) +
    geom_density() +
    labs(title = paste0("Density plot of ", xvar),
         subtitle = sub) +
    r_ladies_theme() 
  
  plot(myplot)
}

purrr::walk(.x = vars, .f = density_fun, df = child_who, sub = sub)
```


# zscorer

The [zscorer](https://nutriverse.validmeasures.org/zscorer/) package is developed by Mark Myatt and Ernest Guevarra. 

> zscorer refers to the results of the WHO Multicentre Growth Reference Study as standard for calculating the z-scores hence it comes packaged with this reference data.

The WHO Multicentre Growth Reference Study combined longitudinal data for children from birth to 24 months and cross-sectional data for children up to 71 months


```{r eval=FALSE}
install.packages("zscorer")
```

The authors note that:

> The function fails messily when secondPart is outside of the range given in the WGS reference (i.e. 45 to 120 cm for height and 0 to 60 months for age). It is up to you to check the ranges of your data.

Because this function is meant for a specific age and height range, I'll filter the data first.

```{r}
zscorer_df <- child_df %>%
  filter(agemos < 60) %>%
  filter(height_length_cm < 120,
         height_length_cm > 45) %>%
  select(id, sex_num, height_length_cm, weight_kg, agemos) %>%
  drop_na() %>%
  as.data.frame() # cannot be type tbl_df

library(zscorer)

# height for age z-score
zscorer_df$haz <- getCohortWGS(data = zscorer_df,
                                 sexObserved = "sex_num",
                                 firstPart = "height_length_cm",
                                 secondPart = "agemos",
                                 index = "hfa")
# weight for age z-score
zscorer_df$waz <- getCohortWGS(data = zscorer_df,
                                 sexObserved = "sex_num",
                                 firstPart = "weight_kg",
                                 secondPart = "agemos",
                                 index = "wfa")
# weight fot height z-score
zscorer_df$whz <- getCohortWGS(data = zscorer_df,
                                 sexObserved = "sex_num",
                                 firstPart = "weight_kg",
                                 secondPart = "height_length_cm",
                                 index = "wfh")

zscorer_df %>%
  select(haz, waz, whz) %>%
  skimr::skim()
```

# childsds package

The [childsds](https://github.com/mvogel78/childsds) package by Mandy Vogel has child reference data from many other countries (WHO, UK, Germany, Italy, China, and more). 

```{r eval=FALSE}
install.packages("chilsds")
```

You can obtain information about the reference data by doing:
```{r}
childsds::cdc.ref
```

We can calculate the BMI z-score and percentile given the reference charts:

```{r}
library(childsds)

childsds_df <- child_df %>%
  filter(agemos >= 24) %>%
  mutate(bmiz = sds(bmi,
                    age = ageyears,
                    sex = sex, male = "Male", female =  "Female",
                    ref = cdc.ref,
                    item = "bmi",
                    type = "SDS"),
         bmipct = round(pnorm(bmiz)*100,2),
         # or use sds function and type = "perc"
         bmipct2 = sds(bmi,
                    age = ageyears,
                    sex = sex, male = "Male", female =  "Female",
                    ref = cdc.ref,
                    item = "bmi",
                    type = "perc")) %>% 
  glimpse() 
```

Plot the values:
```{r}
sub <- "Using CDC growth charts from childsds package"
vars <- c("bmiz", "bmipct")
  
purrr::walk(.x = vars, .f = density_fun, df = childsds_df, sub = sub)
```

# Other work

There is some other publicly available work in this research area:

* [SITAR](https://github.com/statist7/sitar): SuperImposition by Translation And Rotation model by Tim Cole and colleagues. The SITAR package provides functions for fitting and plotting growth curve models. More background information can be found [here](SuperImposition by Translation And Rotation).

* [Canadian Pediatric Endocrine Group](https://cpeg-gcep.net/content/who-macro-files-cpeg-revision): This group has made available SAS, R, an Stata programs for the WHO growth charts, an R script for [waist circumference z-scores](https://cpeg-gcep.net/content/waist-circumference-and-waist-height-ratio-charts) (NHANES, cycle III), and an R script for the preterm z-scores (Fenton reference). They have also published many [shiny apps](https://cpeg-gcep.net/content/shiny-apps)!


***

Please comment if you have found something that I'm missing. I will write a second post with thoughts on what to do next with our unconf project, mchtoolbox.


[^1]: Thanks [Jenny Draper](https://twitter.com/learithe), [Jennifer Thompson](https://twitter.com/jent103), [Kyle Hamilton](https://twitter.com/kylehamilton), and [Charles Gray](https://twitter.com/cantabile) for being on this team!

[^2]: Thanks [Mara](https://maraaverick.rbind.io/2018/06/meta-sods18-a-call-to-curation/) for the encouragement!




